---
title: "Ruffley & Lutz et al. — statistics computed for the NEE revision"
subtitle: "Flowering time × WUE (δ¹³C): correlations, slopes, covariances — plus the
  Table S28 G-matrix constraint recompute"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
mainfont: Times New Roman
monofont: Menlo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 110)
options(stringsAsFactors = FALSE)
set.seed(1)
```

Everything below is recomputed from the deposited data. Written to answer
Reviewer 2 comment 1 ("no p-value or r² values are listed") and the related
Reviewer 1 points on Fig. 3B/3C.

The final section (*Table S28 — evolvability, respondability, and multivariate
constraint*) was merged in on 2026-08-12 from
`stats-recompute/TableS28_evolvability_explainer.Rmd` (Moi + Claude, 2026-08-11/12):
it documents the G-matrix construction bug behind Tables S28/S29 and the
recomputed constraint statistics. The regeneration pipeline for the supplement
lives in `regenerated-tables/regenerate_S21_S28_S29.R`.

# Source data

```{r paths}
SUB  <- "~/projects/natvar/NatEcoEvo_SubmissionNovember"
DAT  <- file.path(SUB, "Athaliana_Natural_Selection_Conflict/data")
CHAM <- "~/projects/natvar/Ruffley_Lutz_et_al_joint_MS/Data_code_Ruffley_Lutz_et_al"

src <- data.frame(
  file = c("deltaC_extended_group_rep3.csv",
           "deltaC_extended_group_rep3_pval.csv",
           "lines_info_annot.csv",
           "atlas1001_phenotypes_matrix_MR.csv",
           "imputed_wt&mut_dc13.csv"),
  where = c(CHAM, CHAM, CHAM, DAT, DAT),
  what = c("Uli's δ¹³C experiment: 49 line means (29 flc-edited, 19 wild-type progenitors, Col-0), 3 reps each. Fig. 3B source.",
           "The 19 edited lines that have a paired wild-type progenitor, with per-line p-values vs that progenitor.",
           "Metadata for 112 edited lines: progenitor accession, FT16 from 1001G, wild-type flowering time and FLC qRT.",
           "1001 Genomes phenotype matrix, MEASURED values (Delta_13C, FT10, FT16).",
           "Imputed δ¹³C for 306 accessions, separate wild-type and flc-mutant predictions."))
knitr::kable(src[, c("file", "what")])
```

```{r load}
d  <- read.csv(file.path(CHAM, "deltaC_extended_group_rep3.csv"))
pr <- read.csv(file.path(CHAM, "deltaC_extended_group_rep3_pval.csv"))
at <- read.csv(file.path(DAT,  "atlas1001_phenotypes_matrix_MR.csv"))
im <- read.csv(file.path(DAT,  "imputed_wt&mut_dc13.csv"))

# groups. dag_mean = days to flowering, 21 C chamber; deltaC_mean_mean = measured d13C
E   <- subset(d, label == "mut")                 # 29 edited lines
W   <- subset(d, label == "wt")                  # 19 wild-type progenitors
WC  <- subset(d, label %in% c("wt", "col"))      # + Col-0  (n = 20; this is the Fig 3B green fit)
at16 <- at[!is.na(at$FT16) & !is.na(at$Delta_13C), ]
at10 <- at[!is.na(at$FT10) & !is.na(at$Delta_13C), ]
c(edited = nrow(E), progenitors = nrow(W), progenitors_col0 = nrow(WC),
  atlas_FT16 = nrow(at16), atlas_FT10 = nrow(at10))
```

Each of the 29 edited lines comes from a **distinct** accession (no pseudo-replication),
and all 29 have δ¹³C measured in triplicate. The 19 are the subset with a paired
progenitor — *not* a measured-vs-imputed distinction.

```{r independence}
firstnum <- function(v) as.numeric(sapply(v, function(x) {
  m <- regmatches(x, regexpr("[0-9]+", x)); if (length(m)) m else NA }))
c(edited_lines = nrow(E), unique_accessions = length(unique(firstnum(E$line))),
  all_measured_3reps = sum(E$deltaC_mean_count == 3),
  paired_subset_of_29 = all(pr$line %in% E$line))
```

# Core statistics

```{r core}
stat <- function(x, y, label) {
  ok <- !is.na(x) & !is.na(y); x <- x[ok]; y <- y[ok]
  ct <- cor.test(x, y); f <- lm(y ~ x)
  data.frame(set = label, n = length(x),
             r = round(unname(ct$estimate), 3),
             p = signif(ct$p.value, 3),
             slope = round(unname(coef(f)[2]), 2),
             slope_SE = round(summary(f)$coef[2, 2], 2),
             cov = round(cov(x, y), 3),
             sd_d13C = round(sd(x), 3),
             sd_FT = round(sd(y), 2))
}

tab <- rbind(
  stat(pr$deltaC_mean_mean, pr$dag_mean,   "19 edited (paired w/ progenitor)"),
  stat(E$deltaC_mean_mean,  E$dag_mean,    "29 edited (all)"),
  stat(W$deltaC_mean_mean,  W$dag_mean,    "19 progenitors"),
  stat(WC$deltaC_mean_mean, WC$dag_mean,   "20 progenitors + Col-0"),
  stat(at16$Delta_13C,      at16$FT16,     "248 natural accessions (FT16)"),
  stat(at10$Delta_13C,      at10$FT10,     "261 natural accessions (FT10)"))
knitr::kable(tab, row.names = FALSE)
```

Flowering time is days-to-flowering at 21 °C for the chamber sets and FT10/FT16 for the
natural accessions, so slopes are comparable in kind but not identical quantities.

`20 progenitors + Col-0` reproduces the `β = 10.44 ± 16.9` reported in the manuscript,
confirming the Fig. 3B wild-type line includes Col-0.

# Formal tests

```{r tests}
fz <- function(r1, r2, n1, n2) {
  z <- (atanh(r1) - atanh(r2)) / sqrt(1/(n1 - 3) + 1/(n2 - 3))
  c(z = round(z, 3), p = round(2 * (1 - pnorm(abs(z))), 4)) }

rbind(
  `edited vs natural (FT16)`  = fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(at16$Delta_13C, at16$FT16), nrow(E), nrow(at16)),
  `edited vs natural (FT10)`  = fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(at10$Delta_13C, at10$FT10), nrow(E), nrow(at10)),
  `edited vs progenitors+Col0`= fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(WC$deltaC_mean_mean, WC$dag_mean), nrow(E), nrow(WC)))
```

```{r slopetest}
b <- rbind(data.frame(x = E$deltaC_mean_mean,  y = E$dag_mean,  g = "edited"),
           data.frame(x = WC$deltaC_mean_mean, y = WC$dag_mean, g = "progenitor"))
round(summary(lm(y ~ x * g, data = b))$coef, 4)   # interaction = slope difference
```

```{r covtest}
cc <- function(s) cov(s$deltaC_mean_mean, s$dag_mean)
obs <- cc(E) - cc(WC)
boot <- replicate(20000, cc(E[sample(nrow(E), replace = TRUE), ]) -
                          cc(WC[sample(nrow(WC), replace = TRUE), ]))
c(observed_difference = round(obs, 3),
  CI_lower = round(quantile(boot, .025), 3), CI_upper = round(quantile(boot, .975), 3))
```

**The variance test is the robust result.** Editing collapses flowering-time variance
~32-fold while leaving δ¹³C variance untouched.

```{r vartest}
rbind(d13C = c(progenitors = var(WC$deltaC_mean_mean), edited = var(E$deltaC_mean_mean)),
      FT   = c(var(WC$dag_mean), var(E$dag_mean)))
var.test(WC$dag_mean, E$dag_mean)
```

Box's M tests the whole covariance matrix, so it rejects here because of the *variance*,
not the covariance — do not cite it as evidence about covariance.

```{r boxm}
S1 <- cov(cbind(E$deltaC_mean_mean, E$dag_mean));  n1 <- nrow(E)
S2 <- cov(cbind(WC$deltaC_mean_mean, WC$dag_mean)); n2 <- nrow(WC); p <- 2; g <- 2
Sp <- ((n1-1)*S1 + (n2-1)*S2) / (n1 + n2 - g)
M  <- (n1+n2-g)*log(det(Sp)) - ((n1-1)*log(det(S1)) + (n2-1)*log(det(S2)))
cst <- 1 - (sum(1/c(n1-1, n2-1)) - 1/(n1+n2-g)) * (2*p^2+3*p-1) / (6*(p+1)*(g-1))
c(chi2 = round(M*cst, 2), df = (g-1)*p*(p+1)/2,
  p = signif(pchisq(M*cst, (g-1)*p*(p+1)/2, lower.tail = FALSE), 3))
```

# Covariance scaling

Scaling each set by its **own** SDs returns the correlation and discards the magnitude
information. Scaling all sets by a **common** reference (the natural population) keeps it.

```{r scaling}
sdx0 <- sd(at16$Delta_13C); sdy0 <- sd(at16$FT16)
sc <- function(x, y, lab) data.frame(set = lab, r = round(cor(x, y), 3),
  cov_raw = round(cov(x, y), 3),
  cov_ownSD = round(cov(x/sd(x), y/sd(y)), 3),
  cov_commonSD = round(cov(x/sdx0, y/sdy0), 3))
knitr::kable(rbind(
  sc(E$deltaC_mean_mean, E$dag_mean,   "29 edited"),
  sc(WC$deltaC_mean_mean, WC$dag_mean, "20 progenitors + Col-0"),
  sc(at16$Delta_13C, at16$FT16,        "248 natural (FT16)")), row.names = FALSE)
```

Do **not** mean-standardise δ¹³C (Houle-style evolvability): its mean is negative on an
arbitrary reference scale, so mean-scaled values are sign-inverted nonsense.

# Relative variation (percentage)

δ¹³C has a negative mean on an arbitrary (VPDB) zero, so a signed CV is negative and
uninterpretable. We therefore report `100 * SD / |mean|` for δ¹³C, and a conventional
CV for flowering time, which has a true zero.

```{r cv}
relsd <- function(x) 100 * sd(x) / abs(mean(x))
cvpct <- function(x) 100 * sd(x) / mean(x)
cvrow <- function(lab, x, y) data.frame(
  set = lab, n = length(x),
  d13C_relSD_pct = round(relsd(x), 2),
  FT_CV_pct = round(cvpct(y), 2))

cvtab <- rbind(
  cvrow("29 edited",              E$deltaC_mean_mean,  E$dag_mean),
  cvrow("20 progenitors + Col-0", WC$deltaC_mean_mean, WC$dag_mean),
  cvrow("19 progenitors",         W$deltaC_mean_mean,  W$dag_mean),
  cvrow("248 natural (FT16)",     at16$Delta_13C,      at16$FT16),
  cvrow("261 natural (FT10)",     at10$Delta_13C,      at10$FT10))
knitr::kable(cvtab, row.names = FALSE)

c(d13C_ratio_progenitors_over_edited = round(relsd(WC$deltaC_mean_mean) /
                                             relsd(E$deltaC_mean_mean), 2),
  FT_ratio_progenitors_over_edited   = round(cvpct(WC$dag_mean) /
                                             cvpct(E$dag_mean), 2))
```

Relative δ¹³C variation is essentially identical between edited lines and progenitors
(ratio ≈ 0.97), while relative flowering-time variation is ~1.8× lower in the edited
lines — the same conclusion as the F tests, without the mean shift that inflates the
raw 32-fold variance ratio.

Only the progenitors-vs-edited comparison is condition-controlled (same chamber, same
experiment). The natural panels were phenotyped at 16 °C and 10 °C, and CV depends
strongly on condition — the same accessions give 31.5% at FT16 and 20.2% at FT10 — so
chamber-vs-atlas CV ratios are not interpretable.

# Measured vs imputed

`imputed_wt&mut_dc13.csv` holds separate wild-type and flc-mutant predictions for 306
accessions. For the edited lines that were actually measured it contains the **measured**
values (fidelity r = 1), so imputation changes nothing there.

```{r imputed}
d$acc <- firstnum(d$line); d$acc[d$label == "col"] <- 6909
m <- merge(d, im, by.x = "acc", by.y = "id", all.x = TRUE)
Ei <- subset(m, label == "mut"); Wi <- subset(m, label == "wt")
knitr::kable(rbind(
  stat(Ei$flc_mut_dc13,     Ei$dag_mean, "edited, imputed"),
  stat(Ei$deltaC_mean_mean, Ei$dag_mean, "edited, measured"),
  stat(Wi$wt_dc13,          Wi$dag_mean, "progenitors, imputed"),
  stat(Wi$deltaC_mean_mean, Wi$dag_mean, "progenitors, measured")), row.names = FALSE)

c(fidelity_edited = round(cor(Ei$flc_mut_dc13, Ei$deltaC_mean_mean, use = "complete"), 3),
  fidelity_wildtype = round(cor(Wi$wt_dc13, Wi$deltaC_mean_mean, use = "complete"), 3))
```

**Scale warning.** The two imputed columns are on different scales — mean wild-type
−35.6‰ vs mean mutant −30.8‰, an offset of ~4.8‰ — which is why the analysis code
divides them by different constants (−37.21 and −31.14). This is a plausible contributor
to the wild-type/mutant non-overlap Reviewer 1 questioned in Fig. 3C.

```{r scaleoffset}
c(mean_wt_imputed = round(mean(im$wt_dc13), 2), sd_wt = round(sd(im$wt_dc13), 3),
  mean_mut_imputed = round(mean(im$flc_mut_dc13), 2), sd_mut = round(sd(im$flc_mut_dc13), 3),
  offset = round(mean(im$wt_dc13) - mean(im$flc_mut_dc13), 2))
```

# Flowering time vs fitness (Reviewer 2, comment 1b)

Reviewer 2 asked for the flowering-time/seed-weight relationship over the same lines as
Fig. 3C. Source is the growth-chamber experiment (`optimal` treatment), not Uli's δ¹³C run.

```{r ftfit}
GC <- paste0("/Users/moisesexpositoalonso/Library/CloudStorage/",
             "GoogleDrive-moiexpositoalonso@berkeley.edu/Shared drives/MOI-LAB/PROJECTS/",
             "field_experiment_ranges/climate.change.evo.exp/field_prep_year1/crispr/growth chamber")
gc <- read.csv(file.path(GC, "wrangled_data_crispr.csv"))
sw <- read.csv(file.path(GC, "seed_weights.csv"))
gd <- merge(gc, sw, by = "NEW_ID")
gd$weight_w_paper <- suppressWarnings(as.numeric(gd$weight_w_paper))
gd$rSeedWeight <- gd$weight_w_paper - 0.3992667          # as in AtlasofPhenotypes.Rmd
gd$founder <- as.character(gd$founder)
gd$founder[is.na(gd$founder) | gd$founder == ""] <- "c"  # c = CRISPR, f = wild-type founder
gopt <- subset(gd, drought == "optimal")

# replicate wrangle_Expdata(): per line, drop NA seed weight, then average
wrangle <- function(dat) do.call(rbind, lapply(unique(dat$real_id), function(i) {
  t <- dat[dat$real_id == i, ]; t <- t[!is.na(t$rSeedWeight), ]
  if (!nrow(t)) return(NULL)
  data.frame(id = i, ft = mean(t$days_to_flowering, na.rm = TRUE),
             fit = mean(t$rSeedWeight, na.rm = TRUE)) }))
WTg  <- wrangle(subset(gopt, founder == "f"))
MUTg <- wrangle(subset(gopt, founder == "c"))

knitr::kable(rbind(stat(WTg$ft,  WTg$fit,  "wild type"),
                   stat(MUTg$ft, MUTg$fit, "CRISPR flc")), row.names = FALSE)
```

```{r ftfit_tests}
bb <- rbind(data.frame(ft = WTg$ft, fit = WTg$fit, g = "wt"),
            data.frame(ft = MUTg$ft, fit = MUTg$fit, g = "mut"))
round(summary(lm(fit ~ ft * g, data = bb))$coef, 5)   # ft:gwt = slope difference

zz <- (atanh(cor(MUTg$ft, MUTg$fit)) - atanh(cor(WTg$ft, WTg$fit))) /
      sqrt(1/(nrow(MUTg) - 3) + 1/(nrow(WTg) - 3))
c(fisher_z = round(zz, 3), p = round(2 * (1 - pnorm(abs(zz))), 4))
```

```{r ftfit_plot, fig.height=4.2}
plot(WTg$ft, WTg$fit, pch = 19, col = "#4E9A6A",
     xlab = "Flowering time (days)", ylab = "Seed weight (g, paper-corrected)",
     main = "Flowering time vs fitness, growth chamber")
points(MUTg$ft, MUTg$fit, pch = 19, col = "#D7473A")
abline(lm(fit ~ ft, data = WTg),  col = "#4E9A6A", lwd = 2)
abline(lm(fit ~ ft, data = MUTg), col = "#D7473A", lwd = 2)
legend("topright", bty = "n", pch = 19, col = c("#4E9A6A", "#D7473A"),
       legend = c("wild type  r = -0.21, p = 0.195", "CRISPR flc  r = +0.25, p = 0.098"))
```

Neither correlation is individually significant, but they differ from each other
(interaction p = 0.033; Fisher z = 2.07, p = 0.038): the relationship reverses sign rather
than disappearing. Over the same lines, fitness vs δ¹³C is r = −0.19 (p = 0.24) in wild
types and r = 0.40 (p = 0.004) in edited lines — so WUE carries the stronger and only
individually significant fitness association in the edited background.

**Caveat.** Flowering time in this chamber experiment spans 27.3–137.5 d in the edited
lines, unlike 13.9–35.8 d in Uli's δ¹³C panel. The two experiments are not the same set of
lines, so the 32-fold variance collapse reported above applies to Uli's panel only.

# Summary of the Fig. 3B/3C tests

| Claim | Test | Result |
|---|---|---|
| Editing changed the correlation | Fisher z vs natural population | p = 0.97 — no |
| Editing changed the slope | interaction term | p = 0.68 — no |
| Editing changed the covariance | bootstrap | p = 0.84 — no |
| Editing collapsed flowering-time variance | F test | **p = 6×10⁻¹⁴ — yes, 32-fold** |
| Wild-type relationship in Fig. 3B is significant | correlation | p = 0.55 — no |

The constraint is relaxed because flowering-time variance collapses, not because the
genetic correlation between the traits changes.


# Table S28 — evolvability, respondability, and multivariate constraint

*Merged from `stats-recompute/TableS28_evolvability_explainer.Rmd` (2026-08-11,
updated 2026-08-12) when that folder was dissolved into
`figures-supp/FigS2-remake/` and this document. The Hansen & Houle framework
behind Table S28 — a walkthrough — plus the provenance and rebuild of the G
matrix behind it.*


```{r s28-palette}
# palette (colorblind-safe, validated)
col_beta <- "#2a78d6"   # selection gradient (blue)
col_dz   <- "#eb6834"   # predicted response (orange)
col_aux  <- "#1baf7a"   # third series (aqua)
col_ell  <- "#898781"   # G ellipse / muted ink
col_ink  <- "#0b0b0b"
```

## The setup: one equation, four numbers

Everything in Table S28 is geometry applied to the multivariate breeder's equation
(Lande 1979):

$$\Delta \bar{z} \;=\; G\,\beta$$

- $\beta$ — the **selection gradient**: the direction (in 12-trait space) in which
  selection pushes the phenotype. From our multivariate selection analysis
  (one $\beta$ per fitness measure: survival, fecundity, composite fitness).
- $G$ — the **additive genetic variance–covariance matrix** of the same 12 traits.
- $\Delta \bar{z}$ — the **predicted response**: where the population mean actually
  moves in one generation.

The key point of Hansen & Houle (2008): selection *proposes* a direction, but $G$
*disposes*. $G$ acts on $\beta$ like a lens — it stretches the response along
directions rich in genetic variance and shrinks it along depleted ones, so
$\Delta \bar z$ generally points somewhere *between* $\beta$ and $g_{max}$ (the
leading eigenvector of $G$).

For a **unit-length** $\beta$ (all formulas below assume $|\beta| = 1$), the
framework defines:

| Quantity | Formula | Meaning |
|---|---|---|
| **Evolvability** | $e(\beta) = \beta^\top G \beta$ | additive genetic variance available *in the direction selection favours* |
| **Respondability** | $r(\beta) = |G\beta|$ | how far the population moves, in any direction |
| **Angle** | $\theta = \angle(\beta, G\beta)$ | how far the response is rotated away from where selection points |
| **Conditional evolvability** | $c(\beta) = \left(\beta^\top G^{-1} \beta\right)^{-1}$ | variance along $\beta$ when *all orthogonal directions are held constant* by stabilising selection |
| **Autonomy** | $a(\beta) = c(\beta)/e(\beta)$ | fraction of the variance along $\beta$ that is free of the rest of the phenotype |

Table S28 reports the first three (plus the cosine of the angle). Conditional
evolvability and autonomy are the rest of the framework, worth knowing.

### The identity that ties them together

$$\cos\theta \;=\; \frac{\beta \cdot G\beta}{|\beta|\,|G\beta|} \;=\; \frac{\beta^\top G \beta}{|G\beta|} \;=\; \frac{e(\beta)}{r(\beta)}$$

So the cosine alignment is **not an independent third line of evidence** — it is
exactly evolvability over respondability. Equivalently, the projection of the
response onto the selection direction has length $r\cos\theta = e$. That gives the
cleanest one-sentence intuition:

> **Respondability is how much the population moves; evolvability is how much of
> that movement is in the direction selection actually pushes. Everything else is
> dragged sideways by genetic covariances.**

Check it against Table S28:

```{r identity-check}
S28 <- data.frame(
  fitness        = c("Survival", "Composite fitness", "Fecundity (seeds)"),
  evolvability   = c(0.29113, 0.28881, 0.21507),
  respondability = c(0.41160, 0.41324, 0.60079),
  angle_deg      = c(44.985, 45.663, 69.024),
  cos_align      = c(0.70729, 0.69888, 0.35798)
)
S28$e_over_r        <- with(S28, evolvability / respondability)
S28$angle_from_cos  <- acos(S28$cos_align) * 180 / pi
knitr::kable(S28, digits = 3,
  caption = "Table S28 values, plus the internal identity: cos θ = e/r, θ = acos(cos θ).")
```

The last two columns reproduce the reported cosine and angle exactly — three
numbers per row, two degrees of freedom. (Internal consistency is necessary but
not sufficient: as shown below, the $G$ behind these published values was not a
valid covariance matrix.)

## Geometric intuition in two traits

A 2-trait toy makes the whole framework visible. The grey ellipse is the 1-SD
contour of $G$; blue is $\beta$ (unit length); orange is $\Delta\bar z = G\beta$.

```{r toy-figure, fig.width=9, fig.height=3.8}
draw_case <- function(G, beta, main) {
  beta <- beta / sqrt(sum(beta^2))
  dz   <- as.vector(G %*% beta)
  e    <- as.numeric(t(beta) %*% G %*% beta)
  r    <- sqrt(sum(dz^2))
  th   <- acos(sum(beta * dz) / r) * 180 / pi
  # 1-SD ellipse of N(0, G)
  E   <- eigen(G)
  tt  <- seq(0, 2 * pi, length.out = 200)
  ell <- E$vectors %*% diag(sqrt(pmax(E$values, 0))) %*% rbind(cos(tt), sin(tt))
  lim <- max(abs(ell), abs(dz), 1) * 1.15
  plot(NA, xlim = c(-lim, lim), ylim = c(-lim, lim), asp = 1,
       axes = FALSE, xlab = "", ylab = "", main = main, font.main = 1, cex.main = 0.95)
  abline(h = 0, v = 0, col = "#e5e4e0")
  lines(ell[1, ], ell[2, ], col = col_ell, lwd = 2)
  # projection of dz onto beta (dashed): its length IS e(beta)
  proj <- sum(dz * beta) * beta
  segments(dz[1], dz[2], proj[1], proj[2], lty = 3, col = col_ell)
  arrows(0, 0, beta[1], beta[2], length = 0.08, lwd = 2.5, col = col_beta)
  arrows(0, 0, dz[1],  dz[2],  length = 0.08, lwd = 2.5, col = col_dz)
  legend("topleft", bty = "n", cex = 0.8, text.col = "#52514e",
         legend = c(sprintf("e = %.2f", e), sprintf("r = %.2f", r),
                    sprintf("θ = %.0f°", th)))
}

Gc <- rbind(c(1, 0.8), c(0.8, 1))   # two traits, r_g = 0.8
op <- par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))
draw_case(Gc, c(1, 1),  "A. β along g_max:\nevolvable AND aligned")
draw_case(Gc, c(1, 0),  "B. β on one trait:\nresponse deflected toward g_max")
draw_case(Gc, c(1, -1), "C. β along g_min:\naligned but barely evolvable")
par(op)
```

The three panels are the three qualitatively different fates of a selection
gradient (blue = $\beta$, orange = $\Delta\bar z$, dotted = projection of the
response back onto $\beta$, whose length is $e$):

- **A — evolvable and aligned.** $\beta$ points along $g_{max}$. The response is
  large and exactly where selection wants it: $e = r$, $\theta = 0$.
- **B — the deflection case (this is our survival/fitness situation).** Selection
  acts on one trait only, but the genetic correlation drags the correlated trait
  along: the response rotates toward $g_{max}$. Note $e < r$: the population
  moves further than the variance along $\beta$ alone would allow — but partly in
  a direction nobody asked for. **With $r_g = 0.8$ between just two traits, the
  deflection is already 39°.** Our fecundity case (69°) is this phenomenon
  compounded across 12 dimensions.
- **C — aligned but not evolvable.** $\beta$ points along $g_{min}$ — antagonistic
  to the correlation (e.g., selection to *increase* one trait while *decreasing*
  a positively correlated one). Because $g_{min}$ is an eigenvector, there is no
  deflection at all ($\theta = 0$) — but $e$ has collapsed to 0.2. **The angle
  alone can therefore look reassuring while evolution is nearly stalled.** You
  need $e$ *and* $\theta$ together, which is why S28 reports both.

## So *when* is something evolvable, and when is it not?

This is the conceptual payoff of the framework. "Evolvable" is not a property of
a trait — it is a property of a **direction** in trait space.

**1. Univariate heritability is almost never the binding constraint.** Every one
of our 12 traits has substantial genetic variance on its own (the diagonal of $G$
— see below). Judged trait-by-trait, everything is evolvable. That criterion is
too weak.

**2. What matters is variance along $\beta$, and it is bounded by the
eigenvalues:** $\lambda_{min} \le e(\beta) \le \lambda_{max}$. If selection
happens to point along a direction where $G$ is thin — a *nearly-null subspace* —
evolvability can approach zero even though every single trait is heritable.
Directions of low variance arise precisely when selection is **antagonistic to
the correlation structure** (panel C): wanting more of one thing and less of a
genetically coupled other. That is the geometry of evolutionary conflict.

**3. The expected evolvability of a random direction is the mean eigenvalue**,
$\bar e = \operatorname{tr}(G)/n$. This gives a natural yardstick: is the
direction selection actually favours richer or poorer in variance than a random
direction would be? (We compute this for our $G$ below — the answer is *poorer*.)

**4. Conditional evolvability can be much smaller than evolvability.** $e(\beta)$
assumes the rest of the phenotype is free to drag along. If the other directions
are under stabilising selection (usually true), the usable variance is
$c(\beta) \le e(\beta)$, sometimes dramatically so when the focal direction is
strongly integrated with the rest. A trait can look evolvable in isolation and be
nearly frozen in context.

**5. "Heritability is not evolvability" (Hansen, Pélabon & Houle 2011).**
$h^2 = V_A/V_P$ measures accuracy of parent–offspring resemblance, not capacity
to change: a trait can have high $h^2$ and trivial usable variance, and vice
versa, because $V_P$ is in the denominator. Hansen & Houle therefore work with
**mean-standardised** variance ($I_A = V_A/\bar z^2$, "percent change per unit
selection"). Keep this in mind when reading our numbers — see the
standardisation caveat at the end.

**In one line: a phenotype is evolvable in the direction of selection when
$e(\beta)$ is a healthy fraction of $\bar e$, and unconstrained when additionally
$\theta$ is small and $c(\beta) \approx e(\beta)$. It is constrained when
selection points into a variance-poor, strongly integrated direction — regardless
of how heritable every individual trait is.**

## The real thing: our 12-trait G matrix

Where the deposited matrix (`data/Gmatrix_imp_multivar_corr_matrix.rda`, behind
published Tables S28/S29) came from: the **diagonal** is each trait's univariate
GEMMA LMM PVE (Table S16); each **off-diagonal** is the genetic covariance
$V_g[2,1]$ from that trait pair's bivariate mvLMM, pasted in from the 66 pair
logs (all 66 verified below). Every entry is a legitimate estimate — the matrix
is *approximately* right — but the assembly is not exactly a covariance matrix:
each pair run implies its own per-trait variances, and those were discarded in
favour of the univariate PVE from a different model. The symptom of the mixed
scales sits at the bottom of published Table S29, **two negative eigenvalues**,
impossible for a true covariance matrix — even though nearly every pairwise
Cauchy–Schwarz bound $|G_{ij}| \le \sqrt{G_{ii}G_{jj}}$ is individually
satisfied. A matrix can pass almost every pairwise sanity check and still be
impossible as a whole.

```{r load-G}
G <- readRDS("data/Gmatrix_imp_multivar_corr_matrix.rda")   # deposited; RDS despite .rda
short <- c("FT16", "Dormancy", "RootGR", "RootAngle", "Delta13C", "StomDens",
           "StomIdx", "VernGrowth", "Germ10C", "GrowthRate", "RGR", "ABA")
rownames(G) <- colnames(G) <- short
round(eigen(G, symmetric = TRUE)$values, 3)   # reproduces Table S29, negatives included
```


### The rebuild: pairwise genetic correlations on a PVE diagonal

**Decision (Moi, 2026-08-12):** treat the univariate PVE as the better truth for
each trait's genetic variance (the univariate GEMMA runs have the larger
per-trait $n$); keep from each pairwise mvLMM only the scale-free genetic
correlation $r_g = V_{g,12}/\sqrt{V_{g,11}\,V_{g,22}}$; and recompute every
covariance from those two ingredients, $G_{ij} = r_{g,ij}\sqrt{PVE_i\,PVE_j}$:

$$G \;=\; D_{PVE}^{1/2}\;\mathrm{nearPD}(R_g)\;D_{PVE}^{1/2}$$

This is the deposited construction's own sourcing (univariate diagonal, mvLMM
off-diagonals) done consistently. The `nearPD` projection is needed because 66
independently estimated pairwise correlations are not jointly PSD (raw $R_g$
eigenvalues run to −0.31; the projection moves no correlation by more than
0.11). `regenerated-tables/regenerate_S21_S28_S29.R` regenerated the supplement
with this $G$; the chunk below rebuilds it live, re-verifying provenance (every
deposited off-diagonal equals its log's $V_g[2,1]$) and the match to the
regenerated file.

```{r final-G}
G_dep  <- readRDS("data/Gmatrix_imp_multivar_corr_matrix.rda")
traits <- rownames(G_dep)                    # full trait names, deposited order
logs <- list.files("multivarGWAS/all_impmGWAS_logoutput", pattern = "log.txt$",
                   full.names = TRUE)
stopifnot(length(logs) == 66)
Rg <- matrix(NA_real_, 12, 12, dimnames = list(traits, traits)); diag(Rg) <- 1
for (f in logs) {
  base <- sub("^imp_mGWAS_", "", sub("\\.log\\.txt$", "", basename(f)))
  t1  <- traits[startsWith(base, paste0(traits, "_"))][1]
  t2  <- substring(base, nchar(t1) + 2)
  ll  <- readLines(f, warn = FALSE)
  i   <- grep("REMLE estimate for Vg", ll)[1]
  v11 <- as.numeric(strsplit(trimws(ll[i + 1]), "\\s+")[[1]][1])
  v2  <- as.numeric(strsplit(trimws(ll[i + 2]), "\\s+")[[1]])
  stopifnot(abs(G_dep[t1, t2] - v2[1]) < 1e-6)   # provenance: log == deposited
  Rg[t1, t2] <- Rg[t2, t1] <- v2[1] / sqrt(v11 * v2[2])
}
Rg_pd   <- as.matrix(Matrix::nearPD(Rg, corr = TRUE)$mat)
Dp      <- diag(sqrt(diag(G)))               # deposited diagonal = univariate PVE
G_final <- Dp %*% Rg_pd %*% Dp
dimnames(G_final) <- list(short, short)

# byte-identical to the regenerated supplement's G
stopifnot(max(abs(G_final - readRDS("regenerated-tables/G_regenerated.rda"))) < 1e-12)
ev_final <- eigen(G_final, symmetric = TRUE)
round(ev_final$values, 4)
```

Constraint geometry with the final $G$ and the deposited Table S13 selection
gradients (Madrid low-water environment, `mli`):

```{r final-angles}
s13 <- readRDS("data/multivariate_SelectionResults_targettraits_1122122.rda")  # S13 betas
map <- c(ABA = "ABA", Delta13C = "Delta13C", Dormancy = "Dormancy",
         FloweringTime = "FT16", GerminationPerc = "Germ10C",
         GrowthRate = "GrowthRate", RGR = "RGR", RootRGR = "RootGR",
         RootHorizIndex = "RootAngle", StomataDensity = "StomDens",
         StomataSize = "StomIdx", Vernalization = "VernGrowth")
get_beta <- function(fit, env = "mli") {
  sub <- s13[s13$env == env & s13$fitness == fit, ]
  b <- setNames(sub$mean.x, map[sub$trait])[short]   # mean.x = the beta column
  b / sqrt(sum(b^2))
}
geom <- function(Gm, b) {
  dz <- as.vector(Gm %*% b)
  e  <- sum(b * dz); r <- sqrt(sum(dz^2))
  c(e = e, r = r, angle = acos(e / r) * 180 / pi)
}
fits  <- c(Survival = "rSurvival_fruit", Fitness = "rFitness", Fecundity = "rSeeds")
betas <- lapply(fits, get_beta)
knitr::kable(round(t(sapply(betas, function(b) geom(G_final, b))), 3), caption =
  "Final G, deposited S13 mli gradients — matches regenerated-tables/S28_regenerated_mli.csv.")
```

**Survival 54.3°, composite fitness 50.9°, fecundity 65.9°** (published S28:
45.0°, 45.7°, 69.0°) — the constraint story survives the rebuild, and these
values reproduce the regenerated Table S28 exactly. Two flags: (1) `nearPD`
lands *on* the PSD boundary, so the two smallest eigenvalues of the final $G$
are exactly zero (rank 10 of 12) — harmless for $e$, $r$, $\theta$, fatal for
$G^{-1}$ (see the conditional section); (2) the true pairwise correlations are
larger than the deposited covariances-read-as-correlations suggested (median
$|r_g|$ 0.36, a quarter of pairs above 0.55, max 0.93).

### Evolvability along every single-trait axis vs. along the eigenvectors

Setting $\beta = (0, \dots, 1, \dots, 0)$ makes $e(\beta) = G_{ii}$: the diagonal
of a variance-standardised $G$ is each trait's own standardised additive variance
— the heritability-scale values, identical in the deposited and final versions
(the diagonal was never the problem). Compare those to the eigenvalue range of
the final $G$:

```{r trait-axes, fig.height=4.2}
e_trait <- diag(G_final)
op <- par(mar = c(4, 7, 1, 1))
bp <- barplot(sort(e_trait), horiz = TRUE, las = 1, col = col_beta, border = NA,
              xlab = "e(β) along each single-trait axis", cex.names = 0.8,
              xlim = c(0, max(ev_final$values) * 1.05))
abline(v = mean(ev_final$values), col = col_ink, lty = 2)
abline(v = max(ev_final$values),  col = col_aux, lty = 2)
text(mean(ev_final$values), max(bp), "mean λ", pos = 4, cex = 0.75, col = col_ink)
text(max(ev_final$values),  max(bp), "λ_max (g1)", pos = 2, cex = 0.75, col = col_aux)
par(op)
```

Five of the twelve trait axes clear the mean eigenvalue (the $h^2 \approx$
0.82–0.94 block: root growth, flowering time, dormancy, growth rate,
germination); the stomatal traits sit near 0.2. But even the lowest trait axis
carries an order of magnitude more variance than the bottom eigendirections of
$G$ ($\lambda_{10} = 0.018$, $\lambda_{11} = \lambda_{12} \approx 0$) — judged
univariately, every trait is evolvable (point 1 above). The interesting question
is what happens along the directions selection actually favours.

### Where do the selection directions sit?

We sample thousands of random unit directions in the 12-dimensional trait space,
compute $e(\beta)$ for each (final $G$), and place the three selection
directions in that distribution.

```{r random-directions, fig.height=4}
set.seed(1)
n <- 20000
X <- matrix(rnorm(n * 12), n, 12)
B <- X / sqrt(rowSums(X^2))                # random unit vectors
e_rand <- rowSums((B %*% G_final) * B)     # beta' G beta, row-wise

obs <- sapply(betas, function(b) geom(G_final, b)["e"])
names(obs) <- names(fits)

op <- par(mar = c(4, 4, 1, 1))
hist(e_rand, breaks = 80, col = "#cde2fb", border = "white",
     main = "", xlab = "e(β) over random directions", ylab = "directions")
abline(v = mean(ev_final$values), col = col_ink, lty = 2)
abline(v = obs, col = col_dz, lwd = 2)
text(mean(ev_final$values), par("usr")[4] * 0.95, "mean λ", pos = 4, cex = 0.8)
text(min(obs), par("usr")[4] * 0.75, "fitness", pos = 2, cex = 0.8, col = col_dz)
text(max(obs), par("usr")[4] * 0.55, "survival, fecundity", pos = 4, cex = 0.8, col = col_dz)
par(op)

round(c(mean_lambda = mean(ev_final$values),
        sapply(obs, function(v) mean(e_rand < v))), 3)
```

All three selection directions fall **below the mean** of the evolvability
distribution — the composite-fitness gradient down in the bottom 3% of random
bearings, survival and fecundity in the bottom sixth to fifth. Selection favours
directions that are variance-poor relative to a random direction through this
trait space. That is multivariate constraint stated as a percentile rather than
an angle — a milder version of the same message the deflection angles carry.
(Note the two measures disagree about *which* direction is worst off: fecundity
has the largest deflection angle, but composite fitness sits in the thinnest
part of the evolvability distribution. An angle measures where the response
*goes*; a percentile measures how much variance the direction *has*.)

### Conditional evolvability and autonomy (the part S28 doesn't show)

$c(\beta)$ needs $G^{-1}$ — and here the final $G$ bites back. `nearPD`
projected the raw $R_g$ *onto the boundary* of the PSD cone, so the final $G$
has two eigenvalues that are exactly zero: it is **singular**, rank 10 in 12
dimensions. The projected matrix asserts two trait combinations with literally
zero genetic variance; every single-trait axis overlaps that null space, so
$c(\beta) \approx 0$ for every trait. Bend the zero eigenvalues up with a floor
before inverting, and the answer tracks the floor, not the data:

```{r conditional}
floors <- c(1e-4, 1e-2, 1e-1)
aut <- sapply(floors, function(fl) {
  v  <- pmax(ev_final$values, fl)                  # eigenvalue floor
  Gi <- ev_final$vectors %*% diag(1 / v) %*% t(ev_final$vectors)
  (1 / diag(Gi)) / diag(G_final)                   # autonomy per trait axis
})
colnames(aut) <- paste0("floor_", floors)
round(aut, 3)
```

Median autonomy 0.001, 0.10, or 0.37 depending on an arbitrary regularisation
constant: **conditional evolvability and autonomy are not estimable from this
$G$.** A joint, rather than pairwise, re-estimation of $G$ would be needed
before any conditional claim can be made ($e$, $r$, and $\theta$ are untouched
by this — they never use $G^{-1}$).

## Reading the constraint results with all of this in hand

Published Table S28 (left) against the regenerated values — final $G$, deposited
S13 `mli` gradients (right; identical to `S28_regenerated_mli.csv`):

| Fitness measure | $e$ | $r$ | $\theta$ | | $e$ | $r$ | $\theta$ |
|---|---|---|---|---|---|---|---|
| Survival | 0.291 | 0.412 | 45.0° | | 0.298 | 0.510 | 54.3° |
| Composite fitness | 0.289 | 0.413 | 45.7° | | 0.178 | 0.283 | 50.9° |
| Fecundity | 0.215 | 0.601 | 69.0° | | 0.325 | 0.797 | 65.9° |

- **Survival / composite fitness** are panel B of the toy figure: usable variance
  along $\beta$, response deflected 51–54°, so ~58–63% of the movement is where
  selection points ($\cos\theta$ = 0.58, 0.63). The published ~45° slightly
  *understated* these deflections; the qualitative reading — real, moderate
  constraint — is unchanged.
- **Fecundity** is panel B sliding toward panel C, and the gap to the other two
  survives the rebuild: 65.9° (published: 69.0°), still by far the worst-aligned.
  Only ~41% of the response is along the selection direction, yet fecundity has
  the *largest* response vector ($r$ = 0.80) — the population moves the most
  while accomplishing the least. A big $r$ with a modest $e$ is not power, it is
  leakage: $G$ takes the fecundity gradient and spends much of it on correlated
  side-effects.
- The constraint is strongest for exactly the fitness component that opposes
  survival in our selection analyses — the geometry of $G$ is entangled with the
  survival–fecundity conflict the paper is about: the fecundity gradient puts
  ~64% of its squared weight on the four variance-poorest eigendirections of
  $G$, versus ~37% for survival (panel C logic).
- And the punchline for intuition: the median pairwise $|r_g|$ is 0.36, with a
  quarter of pairs above 0.55 — no single pair, on its own, explains a 51–66°
  deflection. **Constraint is a property of the whole matrix, not of any pair**
  — a dozen moderate-to-strong correlations compound. Judging constraint from a
  correlation table understates it. (And beware the reverse error the deposited
  file invited: its off-diagonals *looked* like mild correlations but were
  covariances — the true $r_g$ are larger.)

## Caveats specific to our numbers

1. **Standardisation.** Hansen & Houle's canonical $e$ is *mean-standardised*
   (interpretable as % change per generation per unit selection). Ours is
   *variance-standardised* (SD scale), because δ¹³C has a negative mean on an
   arbitrary reference scale — mean-scaling it is meaningless. Consequence: our
   $e$ values are comparable *within* this analysis (across the three fitness
   measures, against the eigenvalue yardstick) but not directly against the
   published distribution of mean-scaled evolvabilities in the Hansen & Houle
   compilations.
2. **The deposited $G$ is not a valid covariance matrix; published S28/S29/S21
   rest on it.** Diagonal (univariate PVE) and off-diagonals (raw pairwise mvLMM
   covariances) come from different models — the two negative eigenvalues are
   the symptom. The supplement has been regenerated with the final
   $G = D_{PVE}^{1/2}\,\mathrm{nearPD}(R_g)\,D_{PVE}^{1/2}$
   (`regenerated-tables/`), and this document uses that same $G$ throughout.
3. **Published S28 is not exactly reproducible from deposited files** (its
   original script was never located), but the regenerated S28 supersedes it and
   this document reproduces the regenerated values exactly. Regenerated S21
   remains provisional: the repo's phenotype CSV is a later export than the one
   behind the published betas ($r \approx 0.89$, see the note in
   `regenerate_S21_S28_S29.R`).
4. **No uncertainty.** All geometric quantities here are point estimates —
   eigenvalues, angles, and especially conditional measures deserve confidence
   intervals (e.g., by resampling the pairwise $r_g$ and $h^2$ estimates), which
   do not exist yet.
5. **Constancy of $G$.** $\Delta\bar z = G\beta$ is a one-generation prediction.
   Over many generations selection itself erodes and rotates $G$, so "constraint"
   here means *bias in the immediate response*, not a permanent wall.

## Cheat sheet

| Symbol | Name | Formula ($|\beta|=1$) | Big means | Small means |
|---|---|---|---|---|
| $e(\beta)$ | evolvability | $\beta^\top G\beta$ | variance available where selection points | selection points into a depleted direction |
| $r(\beta)$ | respondability | $|G\beta|$ | population will move a lot (somewhere) | little response in any direction |
| $\theta$ | deflection angle | $\angle(\beta, G\beta)$ | response dragged off-target by covariances | response goes where selection points |
| $\cos\theta$ | alignment | $e/r$ (identity!) | — | — |
| $c(\beta)$ | conditional evolvability | $(\beta^\top G^{-1}\beta)^{-1}$ | direction can respond even with the rest of the phenotype pinned | variance is mortgaged to other traits |
| $a(\beta)$ | autonomy | $c/e$ | direction is quasi-independent | direction is deeply integrated |
| $\bar e$ | average evolvability | $\operatorname{tr}(G)/n$ | (yardstick: $e$ of a random direction) | |

**References.** Lande 1979 ([doi:10.1111/j.1558-5646.1979.tb04678.x](https://doi.org/10.1111/j.1558-5646.1979.tb04678.x));
Hansen & Houle 2008, *J. Evol. Biol.* ([doi:10.1111/j.1420-9101.2008.01573.x](https://doi.org/10.1111/j.1420-9101.2008.01573.x));
Hansen, Pélabon & Houle 2011, "Heritability is not evolvability" ([doi:10.1007/s11692-011-9127-6](https://doi.org/10.1007/s11692-011-9127-6));
Agrawal & Stinchcombe 2009 on constraint from $G$ ([doi:10.1098/rspb.2008.1855](https://doi.org/10.1098/rspb.2008.1855)).
