---
title: "Ruffley & Lutz et al. — statistics computed for the NEE revision"
subtitle: "Flowering time × WUE (δ¹³C): correlations, slopes, covariances"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, dpi = 110)
options(stringsAsFactors = FALSE)
set.seed(1)
```

Everything below is recomputed from the deposited data. Written to answer
Reviewer 2 comment 1 ("no p-value or r² values are listed") and the related
Reviewer 1 points on Fig. 3B/3C.

# Source data

```{r paths}
SUB  <- "~/projects/natvar/NatEcoEvo_SubmissionNovember"
DAT  <- file.path(SUB, "Athaliana_Natural_Selection_Conflict/data")
CHAM <- "~/projects/natvar/Ruffley_Lutz_et_al_joint_MS/Data_code_Ruffley_Lutz_et_al"

src <- data.frame(
  file = c("deltaC_extended_group_rep3.csv",
           "deltaC_extended_group_rep3_pval.csv",
           "lines_info_annot.csv",
           "atlas1001_phenotypes_matrix_MR.csv",
           "imputed_wt&mut_dc13.csv"),
  where = c(CHAM, CHAM, CHAM, DAT, DAT),
  what = c("Uli's δ¹³C experiment: 49 line means (29 flc-edited, 19 wild-type progenitors, Col-0), 3 reps each. Fig. 3B source.",
           "The 19 edited lines that have a paired wild-type progenitor, with per-line p-values vs that progenitor.",
           "Metadata for 112 edited lines: progenitor accession, FT16 from 1001G, wild-type flowering time and FLC qRT.",
           "1001 Genomes phenotype matrix, MEASURED values (Delta_13C, FT10, FT16).",
           "Imputed δ¹³C for 306 accessions, separate wild-type and flc-mutant predictions."))
knitr::kable(src[, c("file", "what")])
```

```{r load}
d  <- read.csv(file.path(CHAM, "deltaC_extended_group_rep3.csv"))
pr <- read.csv(file.path(CHAM, "deltaC_extended_group_rep3_pval.csv"))
at <- read.csv(file.path(DAT,  "atlas1001_phenotypes_matrix_MR.csv"))
im <- read.csv(file.path(DAT,  "imputed_wt&mut_dc13.csv"))

# groups. dag_mean = days to flowering, 21 C chamber; deltaC_mean_mean = measured d13C
E   <- subset(d, label == "mut")                 # 29 edited lines
W   <- subset(d, label == "wt")                  # 19 wild-type progenitors
WC  <- subset(d, label %in% c("wt", "col"))      # + Col-0  (n = 20; this is the Fig 3B green fit)
at16 <- at[!is.na(at$FT16) & !is.na(at$Delta_13C), ]
at10 <- at[!is.na(at$FT10) & !is.na(at$Delta_13C), ]
c(edited = nrow(E), progenitors = nrow(W), progenitors_col0 = nrow(WC),
  atlas_FT16 = nrow(at16), atlas_FT10 = nrow(at10))
```

Each of the 29 edited lines comes from a **distinct** accession (no pseudo-replication),
and all 29 have δ¹³C measured in triplicate. The 19 are the subset with a paired
progenitor — *not* a measured-vs-imputed distinction.

```{r independence}
firstnum <- function(v) as.numeric(sapply(v, function(x) {
  m <- regmatches(x, regexpr("[0-9]+", x)); if (length(m)) m else NA }))
c(edited_lines = nrow(E), unique_accessions = length(unique(firstnum(E$line))),
  all_measured_3reps = sum(E$deltaC_mean_count == 3),
  paired_subset_of_29 = all(pr$line %in% E$line))
```

# Core statistics

```{r core}
stat <- function(x, y, label) {
  ok <- !is.na(x) & !is.na(y); x <- x[ok]; y <- y[ok]
  ct <- cor.test(x, y); f <- lm(y ~ x)
  data.frame(set = label, n = length(x),
             r = round(unname(ct$estimate), 3),
             p = signif(ct$p.value, 3),
             slope = round(unname(coef(f)[2]), 2),
             slope_SE = round(summary(f)$coef[2, 2], 2),
             cov = round(cov(x, y), 3),
             sd_d13C = round(sd(x), 3),
             sd_FT = round(sd(y), 2))
}

tab <- rbind(
  stat(pr$deltaC_mean_mean, pr$dag_mean,   "19 edited (paired w/ progenitor)"),
  stat(E$deltaC_mean_mean,  E$dag_mean,    "29 edited (all)"),
  stat(W$deltaC_mean_mean,  W$dag_mean,    "19 progenitors"),
  stat(WC$deltaC_mean_mean, WC$dag_mean,   "20 progenitors + Col-0"),
  stat(at16$Delta_13C,      at16$FT16,     "248 natural accessions (FT16)"),
  stat(at10$Delta_13C,      at10$FT10,     "261 natural accessions (FT10)"))
knitr::kable(tab, row.names = FALSE)
```

Flowering time is days-to-flowering at 21 °C for the chamber sets and FT10/FT16 for the
natural accessions, so slopes are comparable in kind but not identical quantities.

`20 progenitors + Col-0` reproduces the `β = 10.44 ± 16.9` reported in the manuscript,
confirming the Fig. 3B wild-type line includes Col-0.

# Visualisation

```{r fig3b, fig.height=4.5}
plot(WC$deltaC_mean_mean, WC$dag_mean, pch = 19, col = "#4E9A6A",
     xlim = range(d$deltaC_mean_mean), ylim = range(d$dag_mean),
     xlab = expression(delta^13*C~"(per mil)"), ylab = "Flowering time (days, 21 C)",
     main = "Fig. 3B recomputed, within-genotype fits")
points(E$deltaC_mean_mean, E$dag_mean, pch = 19, col = "#D7473A")
abline(lm(dag_mean ~ deltaC_mean_mean, data = WC), col = "#4E9A6A", lwd = 2)
abline(lm(dag_mean ~ deltaC_mean_mean, data = E),  col = "#D7473A", lwd = 2)
legend("topleft", bty = "n", pch = 19, col = c("#4E9A6A", "#D7473A"),
       legend = c(sprintf("wild type + Col-0: r = %.2f, p = %.2f",
                          cor(WC$deltaC_mean_mean, WC$dag_mean),
                          cor.test(WC$deltaC_mean_mean, WC$dag_mean)$p.value),
                  sprintf("edited (flc): r = %.2f, p = %.3f",
                          cor(E$deltaC_mean_mean, E$dag_mean),
                          cor.test(E$deltaC_mean_mean, E$dag_mean)$p.value)))
```

```{r background}
plot(at16$Delta_13C, at16$FT16, pch = 19, cex = .45, col = "grey55",
     xlab = expression(delta^13*C~"(per mil)"), ylab = "FT16 (days)",
     main = expression("Natural accessions, measured "*delta^13*C*" (n = 248)"))
abline(lm(FT16 ~ Delta_13C, data = at16), col = "black", lwd = 2)
legend("topleft", bty = "n",
       legend = sprintf("r = %.2f, p = %.1e", cor(at16$Delta_13C, at16$FT16),
                        cor.test(at16$Delta_13C, at16$FT16)$p.value))
```

The natural δ¹³C range (−38.7 to −30.8‰) barely overlaps the chamber range
(−31.4 to −29.4‰); the two cannot be overlaid on a raw axis.

# Formal tests

```{r tests}
fz <- function(r1, r2, n1, n2) {
  z <- (atanh(r1) - atanh(r2)) / sqrt(1/(n1 - 3) + 1/(n2 - 3))
  c(z = round(z, 3), p = round(2 * (1 - pnorm(abs(z))), 4)) }

rbind(
  `edited vs natural (FT16)`  = fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(at16$Delta_13C, at16$FT16), nrow(E), nrow(at16)),
  `edited vs natural (FT10)`  = fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(at10$Delta_13C, at10$FT10), nrow(E), nrow(at10)),
  `edited vs progenitors+Col0`= fz(cor(E$deltaC_mean_mean, E$dag_mean),
                                   cor(WC$deltaC_mean_mean, WC$dag_mean), nrow(E), nrow(WC)))
```

```{r slopetest}
b <- rbind(data.frame(x = E$deltaC_mean_mean,  y = E$dag_mean,  g = "edited"),
           data.frame(x = WC$deltaC_mean_mean, y = WC$dag_mean, g = "progenitor"))
round(summary(lm(y ~ x * g, data = b))$coef, 4)   # interaction = slope difference
```

```{r covtest}
cc <- function(s) cov(s$deltaC_mean_mean, s$dag_mean)
obs <- cc(E) - cc(WC)
boot <- replicate(20000, cc(E[sample(nrow(E), replace = TRUE), ]) -
                          cc(WC[sample(nrow(WC), replace = TRUE), ]))
c(observed_difference = round(obs, 3),
  CI_lower = round(quantile(boot, .025), 3), CI_upper = round(quantile(boot, .975), 3))
```

**The variance test is the robust result.** Editing collapses flowering-time variance
~32-fold while leaving δ¹³C variance untouched.

```{r vartest}
rbind(d13C = c(progenitors = var(WC$deltaC_mean_mean), edited = var(E$deltaC_mean_mean)),
      FT   = c(var(WC$dag_mean), var(E$dag_mean)))
var.test(WC$dag_mean, E$dag_mean)
```

Box's M tests the whole covariance matrix, so it rejects here because of the *variance*,
not the covariance — do not cite it as evidence about covariance.

```{r boxm}
S1 <- cov(cbind(E$deltaC_mean_mean, E$dag_mean));  n1 <- nrow(E)
S2 <- cov(cbind(WC$deltaC_mean_mean, WC$dag_mean)); n2 <- nrow(WC); p <- 2; g <- 2
Sp <- ((n1-1)*S1 + (n2-1)*S2) / (n1 + n2 - g)
M  <- (n1+n2-g)*log(det(Sp)) - ((n1-1)*log(det(S1)) + (n2-1)*log(det(S2)))
cst <- 1 - (sum(1/c(n1-1, n2-1)) - 1/(n1+n2-g)) * (2*p^2+3*p-1) / (6*(p+1)*(g-1))
c(chi2 = round(M*cst, 2), df = (g-1)*p*(p+1)/2,
  p = signif(pchisq(M*cst, (g-1)*p*(p+1)/2, lower.tail = FALSE), 3))
```

# Covariance scaling

Scaling each set by its **own** SDs returns the correlation and discards the magnitude
information. Scaling all sets by a **common** reference (the natural population) keeps it.

```{r scaling}
sdx0 <- sd(at16$Delta_13C); sdy0 <- sd(at16$FT16)
sc <- function(x, y, lab) data.frame(set = lab, r = round(cor(x, y), 3),
  cov_raw = round(cov(x, y), 3),
  cov_ownSD = round(cov(x/sd(x), y/sd(y)), 3),
  cov_commonSD = round(cov(x/sdx0, y/sdy0), 3))
knitr::kable(rbind(
  sc(E$deltaC_mean_mean, E$dag_mean,   "29 edited"),
  sc(WC$deltaC_mean_mean, WC$dag_mean, "20 progenitors + Col-0"),
  sc(at16$Delta_13C, at16$FT16,        "248 natural (FT16)")), row.names = FALSE)
```

Do **not** mean-standardise δ¹³C (Houle-style evolvability): its mean is negative on an
arbitrary reference scale, so mean-scaled values are sign-inverted nonsense.

# Relative variation (percentage)

δ¹³C has a negative mean on an arbitrary (VPDB) zero, so a signed CV is negative and
uninterpretable. We therefore report `100 * SD / |mean|` for δ¹³C, and a conventional
CV for flowering time, which has a true zero.

```{r cv}
relsd <- function(x) 100 * sd(x) / abs(mean(x))
cvpct <- function(x) 100 * sd(x) / mean(x)
cvrow <- function(lab, x, y) data.frame(
  set = lab, n = length(x),
  d13C_relSD_pct = round(relsd(x), 2),
  FT_CV_pct = round(cvpct(y), 2))

cvtab <- rbind(
  cvrow("29 edited",              E$deltaC_mean_mean,  E$dag_mean),
  cvrow("20 progenitors + Col-0", WC$deltaC_mean_mean, WC$dag_mean),
  cvrow("19 progenitors",         W$deltaC_mean_mean,  W$dag_mean),
  cvrow("248 natural (FT16)",     at16$Delta_13C,      at16$FT16),
  cvrow("261 natural (FT10)",     at10$Delta_13C,      at10$FT10))
knitr::kable(cvtab, row.names = FALSE)

c(d13C_ratio_progenitors_over_edited = round(relsd(WC$deltaC_mean_mean) /
                                             relsd(E$deltaC_mean_mean), 2),
  FT_ratio_progenitors_over_edited   = round(cvpct(WC$dag_mean) /
                                             cvpct(E$dag_mean), 2))
```

Relative δ¹³C variation is essentially identical between edited lines and progenitors
(ratio ≈ 0.97), while relative flowering-time variation is ~1.8× lower in the edited
lines — the same conclusion as the F tests, without the mean shift that inflates the
raw 32-fold variance ratio.

Only the progenitors-vs-edited comparison is condition-controlled (same chamber, same
experiment). The natural panels were phenotyped at 16 °C and 10 °C, and CV depends
strongly on condition — the same accessions give 31.5% at FT16 and 20.2% at FT10 — so
chamber-vs-atlas CV ratios are not interpretable.

# Measured vs imputed

`imputed_wt&mut_dc13.csv` holds separate wild-type and flc-mutant predictions for 306
accessions. For the edited lines that were actually measured it contains the **measured**
values (fidelity r = 1), so imputation changes nothing there.

```{r imputed}
d$acc <- firstnum(d$line); d$acc[d$label == "col"] <- 6909
m <- merge(d, im, by.x = "acc", by.y = "id", all.x = TRUE)
Ei <- subset(m, label == "mut"); Wi <- subset(m, label == "wt")
knitr::kable(rbind(
  stat(Ei$flc_mut_dc13,     Ei$dag_mean, "edited, imputed"),
  stat(Ei$deltaC_mean_mean, Ei$dag_mean, "edited, measured"),
  stat(Wi$wt_dc13,          Wi$dag_mean, "progenitors, imputed"),
  stat(Wi$deltaC_mean_mean, Wi$dag_mean, "progenitors, measured")), row.names = FALSE)

c(fidelity_edited = round(cor(Ei$flc_mut_dc13, Ei$deltaC_mean_mean, use = "complete"), 3),
  fidelity_wildtype = round(cor(Wi$wt_dc13, Wi$deltaC_mean_mean, use = "complete"), 3))
```

**Scale warning.** The two imputed columns are on different scales — mean wild-type
−35.6‰ vs mean mutant −30.8‰, an offset of ~4.8‰ — which is why the analysis code
divides them by different constants (−37.21 and −31.14). This is a plausible contributor
to the wild-type/mutant non-overlap Reviewer 1 questioned in Fig. 3C.

```{r scaleoffset}
c(mean_wt_imputed = round(mean(im$wt_dc13), 2), sd_wt = round(sd(im$wt_dc13), 3),
  mean_mut_imputed = round(mean(im$flc_mut_dc13), 2), sd_mut = round(sd(im$flc_mut_dc13), 3),
  offset = round(mean(im$wt_dc13) - mean(im$flc_mut_dc13), 2))
```

# Flowering time vs fitness (Reviewer 2, comment 1b)

Reviewer 2 asked for the flowering-time/seed-weight relationship over the same lines as
Fig. 3C. Source is the growth-chamber experiment (`optimal` treatment), not Uli's δ¹³C run.

```{r ftfit}
GC <- paste0("/Users/moisesexpositoalonso/Library/CloudStorage/",
             "GoogleDrive-moiexpositoalonso@berkeley.edu/Shared drives/MOI-LAB/PROJECTS/",
             "field_experiment_ranges/climate.change.evo.exp/field_prep_year1/crispr/growth chamber")
gc <- read.csv(file.path(GC, "wrangled_data_crispr.csv"))
sw <- read.csv(file.path(GC, "seed_weights.csv"))
gd <- merge(gc, sw, by = "NEW_ID")
gd$weight_w_paper <- suppressWarnings(as.numeric(gd$weight_w_paper))
gd$rSeedWeight <- gd$weight_w_paper - 0.3992667          # as in AtlasofPhenotypes.Rmd
gd$founder <- as.character(gd$founder)
gd$founder[is.na(gd$founder) | gd$founder == ""] <- "c"  # c = CRISPR, f = wild-type founder
gopt <- subset(gd, drought == "optimal")

# replicate wrangle_Expdata(): per line, drop NA seed weight, then average
wrangle <- function(dat) do.call(rbind, lapply(unique(dat$real_id), function(i) {
  t <- dat[dat$real_id == i, ]; t <- t[!is.na(t$rSeedWeight), ]
  if (!nrow(t)) return(NULL)
  data.frame(id = i, ft = mean(t$days_to_flowering, na.rm = TRUE),
             fit = mean(t$rSeedWeight, na.rm = TRUE)) }))
WTg  <- wrangle(subset(gopt, founder == "f"))
MUTg <- wrangle(subset(gopt, founder == "c"))

knitr::kable(rbind(stat(WTg$ft,  WTg$fit,  "wild type"),
                   stat(MUTg$ft, MUTg$fit, "CRISPR flc")), row.names = FALSE)
```

```{r ftfit_tests}
bb <- rbind(data.frame(ft = WTg$ft, fit = WTg$fit, g = "wt"),
            data.frame(ft = MUTg$ft, fit = MUTg$fit, g = "mut"))
round(summary(lm(fit ~ ft * g, data = bb))$coef, 5)   # ft:gwt = slope difference

nWT <- sum(complete.cases(WTg$ft, WTg$fit))   # 2 wt lines lack flowering time: n = 39
zz <- (atanh(cor(MUTg$ft, MUTg$fit, use = "complete")) -
       atanh(cor(WTg$ft,  WTg$fit,  use = "complete"))) /
      sqrt(1/(nrow(MUTg) - 3) + 1/(nWT - 3))
c(fisher_z = round(zz, 3), p = round(2 * (1 - pnorm(abs(zz))), 4))
```

```{r ftfit_plot, fig.height=4.2}
plot(WTg$ft, WTg$fit, pch = 19, col = "#4E9A6A",
     xlab = "Flowering time (days)", ylab = "Seed weight (g, paper-corrected)",
     main = "Flowering time vs fitness, growth chamber")
points(MUTg$ft, MUTg$fit, pch = 19, col = "#D7473A")
abline(lm(fit ~ ft, data = WTg),  col = "#4E9A6A", lwd = 2)
abline(lm(fit ~ ft, data = MUTg), col = "#D7473A", lwd = 2)
legend("topright", bty = "n", pch = 19, col = c("#4E9A6A", "#D7473A"),
       legend = c("wild type  r = -0.21, p = 0.195", "CRISPR flc  r = +0.25, p = 0.098"))
```

Neither correlation is individually significant, but they differ from each other
(interaction p = 0.033; Fisher z = 2.08, p = 0.037 at full precision — the published
"z = 2.07, p = 0.038" is the same test computed from the rounded correlations −0.21 and
0.25): the relationship reverses sign rather than disappearing. Over the same lines, fitness vs δ¹³C is r = −0.19 (p = 0.24) in wild
types and r = 0.40 (p = 0.004) in edited lines — so WUE carries the stronger and only
individually significant fitness association in the edited background.

**Caveat.** Flowering time in this chamber experiment spans 27.3–137.5 d in the edited
lines, unlike 13.9–35.8 d in Uli's δ¹³C panel. The two experiments are not the same set of
lines, so the 32-fold variance collapse reported above applies to Uli's panel only.

# Supplement §VIII.2 — every number, in order

One chunk per claim, in the order the supplement makes them, so this section can be
checked line-by-line against the text. Expected values from the current supplement are
noted inline; everything reuses objects loaded above.

## δ¹³C rank consistency between wild types and their flc counterparts

Expected: Spearman ρ = 0.66, p = 0.002; Pearson r = 0.77, p = 1×10⁻⁴, over the 19
wt–flc pairs.

```{r viii2_pairs}
Ep <- subset(d, label == "mut"); Wp <- subset(d, label == "wt")
Ep$acc <- firstnum(Ep$line); Wp$acc <- firstnum(Wp$line)
pw <- merge(Wp[, c("acc", "deltaC_mean_mean")], Ep[, c("acc", "deltaC_mean_mean")],
            by = "acc", suffixes = c("_wt", "_mut"))
c(n_pairs = nrow(pw))
suppressWarnings(   # ties -> approximate Spearman p
  cor.test(pw$deltaC_mean_mean_wt, pw$deltaC_mean_mean_mut, method = "spearman"))
cor.test(pw$deltaC_mean_mean_wt, pw$deltaC_mean_mean_mut)
```

## Measured phenotypes: r, slope, covariance, SD, CV per set

Expected: 19 paired edited r = 0.50, p = 0.030; 29 edited r = 0.38, p = 0.045,
slope 4.58 ± 2.18, cov 0.90, SD 0.44 ‰ / 5.4 d, CV 1.44 % / 24.5 %; 20 progenitors +
Col-0 r = 0.14, p = 0.55, slope 10.44 ± 16.94, cov 1.88, SD 0.42 / 30.8, CV 1.40 / 43.6;
248 natural (FT16) r = 0.38, p = 5.2×10⁻¹⁰, slope 8.63 ± 1.33, cov 12.26, SD 1.19 / 27.0,
CV 3.35 / 31.5. (FT10 gives the same qualitative picture — Core statistics table above.)

```{r viii2_measured}
statcv <- function(x, y, lab) {
  ok <- !is.na(x) & !is.na(y); x <- x[ok]; y <- y[ok]
  cbind(stat(x, y, lab),
        CV_d13C_pct = round(relsd(x), 2), CV_FT_pct = round(cvpct(y), 2)) }

knitr::kable(rbind(
  statcv(pr$deltaC_mean_mean, pr$dag_mean,   "19 edited (paired)"),
  statcv(E$deltaC_mean_mean,  E$dag_mean,    "29 edited"),
  statcv(WC$deltaC_mean_mean, WC$dag_mean,   "20 progenitors + Col-0"),
  statcv(at16$Delta_13C,      at16$FT16,     "248 natural (FT16)")), row.names = FALSE)
```

The Col-0 scale anchor (expected: −31.14 ‰ measured here vs −37.21 ‰ reference used with
the published dataset — ranks comparable across experiments, absolute values not):

```{r viii2_col0}
c(col0_measured_here = d$deltaC_mean_mean[d$label == "col"], col0_published_ref = -37.21)
```

## Fisher z on the measured correlations

Expected: 29 edited vs 20 progenitors + Col-0 z = 0.802, p = 0.423; 29 edited vs 248
natural z = −0.033, p = 0.974.

```{r viii2_fisherz}
rbind(
  `29 edited vs 20 progenitors+Col-0` = fz(cor(E$deltaC_mean_mean, E$dag_mean),
      cor(WC$deltaC_mean_mean, WC$dag_mean), nrow(E), nrow(WC)),
  `29 edited vs 248 natural (FT16)`   = fz(cor(E$deltaC_mean_mean, E$dag_mean),
      cor(at16$Delta_13C, at16$FT16), nrow(E), nrow(at16)))
```

## Variance F tests, measured

Expected: δ¹³C progenitors-vs-edited F = 0.91, p = 0.85; flowering time F = 32.35,
p = 6.4×10⁻¹⁴; vs the 248 natural accessions δ¹³C F = 7.22, p = 6.8×10⁻⁸ and flowering
time F = 24.82, p = 8.4×10⁻¹⁵. The relative-variation claims (δ¹³C CV 1.44 % vs 1.40 %,
flowering-time CV 1.8-fold lower) are the CV columns above and the ratio pair in the
Relative variation section.

```{r viii2_ftests}
vt <- function(x, y, lab) { v <- var.test(x, y)
  data.frame(comparison = lab, F = round(unname(v$statistic), 2),
             df = paste(v$parameter, collapse = ", "), p = signif(v$p.value, 2)) }
knitr::kable(rbind(
  vt(WC$deltaC_mean_mean, E$deltaC_mean_mean, "d13C: 20 progenitors+Col-0 vs 29 edited"),
  vt(WC$dag_mean,         E$dag_mean,         "FT:   20 progenitors+Col-0 vs 29 edited"),
  vt(at16$Delta_13C,      E$deltaC_mean_mean, "d13C: 248 natural vs 29 edited"),
  vt(at16$FT16,           E$dag_mean,         "FT:   248 natural vs 29 edited")),
  row.names = FALSE)
```

## Imputed phenotypes: the three quoted blocks

Expected: 28 edited (CS77069 absent from the imputed panel) r = 0.38, p = 0.049,
slope 4.58 ± 2.22, cov 0.94, SD 0.45 / 5.5, CV 1.47 / 25.0; 19 progenitors (Col-0 absent)
r = 0.14, p = 0.57, slope 2.56 ± 4.46, cov 6.29, SD 1.57 / 29.1, CV 4.43 / 39.7;
1135 natural accessions, imputed δ¹³C × imputed FT16, r = 0.59, p = 8.0×10⁻¹⁰⁶,
slope 19.90 ± 0.82, cov 10.47, SD 0.73 / 24.6, CV 2.02 / 32.5.

```{r viii2_imputed}
c(absent_edited = subset(m, label == "mut" & is.na(flc_mut_dc13))$line,
  absent_wt     = subset(m, label %in% c("wt", "col") & is.na(wt_dc13))$line)
E28 <- subset(m, label == "mut" & !is.na(flc_mut_dc13))              # n = 28
W19 <- subset(m, label %in% c("wt", "col") & !is.na(wt_dc13))        # n = 19
ai  <- read.csv(file.path(DAT, "atlas1001_phenotype_matrix_imputed_withID.csv"))
knitr::kable(rbind(
  statcv(E28$flc_mut_dc13, E28$dag_mean, "28 edited, imputed d13C"),
  statcv(W19$wt_dc13,      W19$dag_mean, "19 progenitors, imputed d13C"),
  statcv(ai$Delta_13C,     ai$FT16,      "1135 natural, imputed (FT16)")),
  row.names = FALSE)
```

## Fisher z and F test on the imputed sets

Expected: 28 edited vs 19 progenitors z = 0.800, p = 0.424; 28 edited vs 1135 natural
z = −1.372, p = 0.170; flowering-time variance F = 27.85, p = 1.3×10⁻¹².

```{r viii2_imputed_tests}
rbind(
  `28 edited vs 19 progenitors (imputed)` = fz(cor(E28$flc_mut_dc13, E28$dag_mean),
      cor(W19$wt_dc13, W19$dag_mean), nrow(E28), nrow(W19)),
  `28 edited vs 1135 natural (imputed)`   = fz(cor(E28$flc_mut_dc13, E28$dag_mean),
      cor(ai$Delta_13C, ai$FT16), nrow(E28), nrow(ai)))
var.test(W19$dag_mean, E28$dag_mean)
```

## Traits vs fitness (seed weight, growth chamber)

Expected: flowering time vs seed weight r = −0.21, p = 0.195, n = 39 in wild types and
r = 0.25, p = 0.098, n = 46 in edited lines; genotype × flowering-time interaction
p = 0.033; Fisher z = 2.07, p = 0.038 (full precision gives 2.08 / 0.037 — the quoted
2.07 is z computed from the rounded correlations). The interaction model is in the
previous section (`ftfit_tests`); `cor.test` drops the two wild-type lines with no
flowering-time record, giving n = 39 of the 41.

```{r viii2_fitness_ft}
unlist(lapply(list(wt = cor.test(WTg$ft, WTg$fit), flc = cor.test(MUTg$ft, MUTg$fit)),
  function(ct) c(n = unname(ct$parameter) + 2, r = round(unname(ct$estimate), 2),
                 p = round(ct$p.value, 3))))
fz(cor(MUTg$ft, MUTg$fit, use = "complete"), cor(WTg$ft, WTg$fit, use = "complete"),
   nrow(MUTg), sum(complete.cases(WTg$ft, WTg$fit)))
```

Expected: fitness vs δ¹³C r = 0.40, p = 0.004 in the edited lines and r = −0.19,
p = 0.24 in the wild types. The two halves use different δ¹³C sources: the edited lines
use chamber-measured `mean_dC13_ko` (the Fig. 3C flc source), while the wild-type value
reproduces exactly from **imputed** `wt_dc13` over the 41 growth-chamber wild-type lines
— which identifies the wild-type half of the published Fig. 3C statistic, listed as
missing in `figures-main/Fig3C-remake/README.md` (the plotted wild-type points remain
PDF-recovered; only the correlation is confirmed here).

```{r viii2_fitness_wue}
fd  <- read.csv(file.path(DAT, "flc_droughtexp_data.csv"))
ik  <- read.csv(file.path(DAT, "imputed_mutant_dC13.csv"))
fmm <- merge(fd, ik[, c("id", "mean_dC13_ko")], by = "id")
cor.test(fmm$mean_dC13_ko, fmm$rSeedWeight)      # edited: expect r = 0.405, p = 0.0047

WTg$acc <- firstnum(as.character(WTg$id))
wwt <- merge(WTg, im, by.x = "acc", by.y = "id")
cor.test(wwt$wt_dc13, wwt$fit)                   # wild type: expect r = -0.19, p = 0.235
```

## Checklist

| §VIII.2 claim | chunk | reproduces? |
|---|---|---|
| ρ = 0.66 / r = 0.77 wt–flc rank consistency | `viii2_pairs` | yes (0.662 / 0.773) |
| measured r, slope, cov, SD, CV — 3 sets + paired 19 | `viii2_measured` | yes |
| Col-0 −31.14 vs −37.21 reference | `viii2_col0` | yes |
| Fisher z 0.802 / −0.033 (measured) | `viii2_fisherz` | yes |
| F tests 0.91, 32.35, 7.22, 24.82 | `viii2_ftests` | yes |
| CV 1.44 vs 1.40 %; FT CV 1.8× lower | `viii2_measured` + Relative variation | yes |
| imputed blocks n = 28 / 19 / 1135 | `viii2_imputed` | yes |
| Fisher z 0.800 / −1.372; F = 27.85 (imputed) | `viii2_imputed_tests` | yes |
| FT–fitness r = −0.21 / 0.25; interaction; z = 2.07 | `viii2_fitness_ft` + `ftfit_tests` | yes (2.08 full precision) |
| δ¹³C–fitness r = −0.19 / 0.40 | `viii2_fitness_wue` | yes |

# Summary

| Claim | Test | Result |
|---|---|---|
| Editing changed the correlation | Fisher z vs natural population | p = 0.97 — no |
| Editing changed the slope | interaction term | p = 0.68 — no |
| Editing changed the covariance | bootstrap | p = 0.84 — no |
| Editing collapsed flowering-time variance | F test | **p = 6×10⁻¹⁴ — yes, 32-fold** |
| Wild-type relationship in Fig. 3B is significant | correlation | p = 0.55 — no |

The constraint is relaxed because flowering-time variance collapses, not because the
genetic correlation between the traits changes.
